---
title: "5.1 Introduction"
author: "Yalin Yang"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
  word_document:
    toc: no
    toc_depth: '3'
---

# 5.1 Introduction

Visualisation is an important tool for insight generation, but it is rare that you get the data in exactly the right form you need. Often you’ll need to create some new variables or summaries, or maybe you just want to rename the variables or reorder the observations in order to make the data a little easier to work with. You’ll learn how to do all that (and more!) in this chapter, which will teach you how to transform your data using the dplyr package and a new dataset on flights departing New York City in 2013.

## 5.1.1 Prerequisites

In this chapter we’re going to focus on how to use the `dplyr` package, another core member of the `tidyverse.` We’ll illustrate the key ideas using data from the nycflights13 package, and use `ggplot2` to help us understand the data.

```{r}
library(nycflights13)
library(tidyverse)
```

## 5.1.2 nycflights13

To explore the basic data manipulation verbs of dplyr, we’ll use `nycflights13::flights`. This data frame contains all 336,776 flights that departed from New York City in 2013. 

```{r}
head(flights)
```

You might also have noticed the row of three (or four) letter abbreviations under the column names. These describe the type of each variable:
* `int` stands for integers.

* `dbl` stands for doubles, or real numbers.

* `chr` stands for character vectors, or strings.

* `dttm` stands for date-times (a date + a time).

There are three other common types of variables that aren’t used in this dataset but you’ll encounter later in the book:

* `lgl` stands for logical, vectors that contain only TRUE or FALSE.

* `fctr` stands for factors, which R uses to represent categorical variables with fixed possible values.

* `date` stands for dates.

## 5.1.3 dplyr basics

In this chapter you are going to learn the five key dplyr functions that allow you to solve the vast majority of your data manipulation challenges:
* Pick observations by their values (`filter()`).
* Reorder the rows (`arrange()`).
* Pick variables by their names (`select()`).
* Create new variables with functions of existing variables (`mutate()`)
* Collapse many values down to a single summary (`summarise()`)


