---
title: "5.3 Arrange rows with `arrange()`"
author: "Yalin Yang"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
  word_document:
    toc: no
    toc_depth: '3'
---

`arrange()` works similarly to `filter() `except that instead of selecting rows, **it changes their order**. It takes a data frame and a set of column names (or more complicated expressions) to order by. If you provide more than one column name, each additional column will be used to break ties in the values of preceding columns:

```{r}
tail(arrange(flights, year, month, day))
```

Use `desc()` to re-order by a column in descending order:

```{r}
tail(arrange(flights, desc(dep_delay)))
```
Missing values are always sorted at the end:

```{r}
df <- tibble(x = c(5, 2, NA))
arrange(df, x)
```

```{r}
arrange(df, desc(x))
```

## 5.3.1 Exercises

### How could you use `arrange()` to sort all missing values to the start? (Hint: use `is.na()`).

```{r}
arrange(df,!is.na(x))
```

### Sort `flights` to find the most delayed flights. Find the flights that left earliest.

```{r}
head(arrange(flights,desc(flights$dep_delay),flights$dep_time))
```

### Sort flights to find the fastest (highest speed) flights.

```{r}
head(arrange(flights, desc(distance / air_time)))
```



