---
title: "5.4 Select columns with `select()`"
author: "Yalin Yang"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
  word_document:
    toc: no
    toc_depth: '3'
---

`select()` allows you to rapidly zoom in on a useful subset using operations based on the names of the variables.

```{r include=FALSE}
# Select columns by name
select(flights, year, month, day)

# Select all columns between year and day (inclusive)
select(flights, year:day)

# Select all columns except those from year to day (inclusive)
select(flights, -(year:day))
```


There are a number of helper functions you can use within `select()`:

*`starts_with("abc")`: matches names that begin with “abc”.

*`ends_with("xyz")`: matches names that end with “xyz”.

*`contains("ijk")`: matches names that contain “ijk”.

*`matches("(.)\\1")`: selects variables that match a regular expression.

* `num_range("x", 1:3)`: matches x1, x2 and x3.

See `?select` for more details.

`select()` can be used to rename variables, but it’s rarely useful because it drops all of the variables not explicitly mentioned. Instead, use `rename()`, which is a variant of `select()` that keeps all the variables that aren’t explicitly mentioned:

```{r}
rename(flights, tail_num = tailnum)
```


Another option is to use `select()` in conjunction with the `everything()` helper. 
This is useful if you have a handful of variables **you’d like to move to the start of the data frame. **

```{r}
select(flights, time_hour, air_time, everything())
```

## 5.4.1 Exercises

Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from flights.
```{r include=FALSE}
select(flights,dep_time,dep_delay,arr_time,arr_delay)
flights[c('dep_time','dep_delay','arr_time','arr_delay')]
select(flights,one_of(c('dep_time','dep_delay','arr_time','arr_delay')))
```

What happens if you include the name of a variable multiple times in a `select()` call?

*Only show one column*
```{r}
head(select(flights,dep_time,dep_time))
```

What does the `one_of()` function do? Why might it be helpful in conjunction with this vector?

```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
head(select(flights,one_of(vars)))
```

Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?

not case sensitive
```{r}
head(select(flights, contains("TIME")))
```

```{r}
head(select(flights, contains("TIME",ignore.case = FALSE)))
```








